Parsing Postfix Log Files

### 2009-04-20 doc/logparser-implementation.tex
Corrections from proofreading for chapters 4 and 5.

This chapter documents the implementation of \parsername{}, based on the
architecture described in \sectionref{parser architecture}.  Any design may
look good when examined in the abstract, but the real test of the design
comes with the first concrete implementation; only then do any practical
difficulties come to light.  Implementing this architecture was relatively
straightforward: the difficulties came not from the architecture and
design, but from anomalies in the log files and Postfix's behaviour, as
described in \sectionref{complications}.  \parsername{} successfully deals
with all the difficulties that were discovered during its development, but
further difficulties may arise during future usage; unfortunately solving
all potential difficulties that may arise is an intractable task, but the
descriptions of the solutions developed thus far (see
\sectionref{complications}) should help if someone needs to develop a
solution of their own.

\parsername{} deals with all the complications of parsing, the
eccentricities and oddities of Postfix log files, presenting the resulting
data in a normalised, simple to use representation.  Unfortunately dealing
with the complications that arise sometimes requires the parser to discard
log lines (e.g.\ \sectionref{aborted delivery attempts}), and, less
frequently, to discard a data structure (e.g.\ \sectionref{timeouts during
data phase}).  \parsername{} can parse log files from Postfix version 2.2
through to version 2.5, and should be capable of parsing log files from
later versions with only minor modifications or an updated ruleset.  The
parser's task is to follow the journey each mail takes through Postfix,
combining the data captured by rules into a coherent whole, saving it in a
useful and consistent form, and performing housekeeping duties.  The
intermingling of log lines from different mails immediately rules out the
possibility of handling each mail in isolation; the parser must handle
multiple mails in parallel, each potentially at a different stage in its
journey, without any interference between mails --- except in the minority
of cases where intra-mail interference is required, e.g.\ mail re-injected
for forwarding (\sectionref{Re-injected mails}).  The best way to deal with
intermingling of log lines is to maintain state information for every
unfinished mail, and to manipulate the appropriate mail correctly for each
log line encountered.

This chapter begins with the assumptions under which \parsername{} was
designed and written, followed by a flowchart showing the most common paths
taken through Postfix and \parsername{}, with a description of the stages
and transitions.  The topic of the next section is the database that
provides storage for the parser: any future work that analyses gathered
data will do so using the database, so the database schema acts as an
\acronym{API}.  A diagram of the database schema is provided, plus
documentation for every table and field.

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users; that
section finishes with the features that are important for this thesis but
not strictly necessary for parsing: the performance data collected by the
framework, the optimisations that can be disabled to show their effect, and
the debugging options it provides.  The second component of the
architecture is the actions: their implementation is documented, with a
graph of how often each action is specified by rules, and a description of
why some actions are more popular than others.  Every action that is part
of \parsername{} is documented, and the actions section concludes with the
process of adding a new action, and how to determine the values for each of
the required fields.  Rules are the final component of the architecture,
and also the most visible to advanced users, e.g.\ systems administrators,
because it is likely that they will need to add their own rules to
recognise their log lines.  An example rule used by \parsername{} is shown,
with every field clearly documented; this is followed by a description of
adding new rules, and how to determine the value of each of the fields.
\parsername{} provides a utility to create regexes from unrecognised log
lines; the algorithm it uses is documented, including the differences
between it and the original algorithm.  The rules section finishes with a
discussion of how \parsername{} uses rule conditions and overlapping rules,
plus a description of the regex snippets provided to ease the process of
writing regexes.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering from a deluge of spam
        or a mail loop.  When parsing individual log files in isolation,
        it is highly likely that some mails will have log lines in previous
        log files, and others will have log lines in subsequent log files;
        to alleviate this problem \parsername{} supports saving and loading
        its state tables, so they can be preserved between log files.

@@

In some ways this task is similar to reverse engineering or replicating a
black box system based solely on its inputs and outputs.  Thus far,
analysis of the log files has been enough to reconstruct Postfix's
behaviour, but for more difficult programs the techniques described in
\cite{black-box-error-reporting} may be useful.  Some advantages come from
treating Postfix as a black box during parser development:

@@

    \item The parser acts as a second source of information about Postfix's
        operation, based on empirical evidence.  A separate project could
        compare the empirical knowledge inherent in \parsername{} with
        Postfix's documentation and source code to see how closely the two
        agree.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; decision boxes and the
difficulties described in \sectionref{complications} are excluded for the
sake of clarity.  The flow chart is intended to be a graphical overview of
how a mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and transitions shown in \figureref{flow chart image} will be
explained later in this section.

@@

    \item This is followed by rejection of \acronym{SMTP} commands by the
        restrictions configured by the administrator
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; failure of the remote client's connection
        \flowchart{DELIVERY\_ERROR}{6}; or some random interleaving of two
        or more of the above.

    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
        normally or with an error.  If Postfix has rejected any
        \acronym{SMTP} commands the resulting data will be saved to the
        database; if there were no rejections there will not be any data to
        save.  Any mails accepted will already have been cloned so their
        data is in another data structure, and will be delivered in the
        same way as mails that entered the system by any other route.

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether by deletion, bouncing, local delivery, or remote delivery.
All four are handled in exactly the same way:

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked with the parent mail \flowchart{TRACK}{9}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

@@

    \item This is the last log line for this particular mail, though it may
        be indirectly referred to if it was re-injected.  The mail is
        cleaned up and entered in the database, then deleted from the state
        tables \flowchart{COMMIT}{11}.

@@

It should be emphasised that the mail delivery process above happens
whether the mail is delivered to a mailbox, piped to a command, delivered
to a remote server, bounced (because of a mail loop, delivery failure, or
five day timeout), or deleted by the administrator, \textit{unless\/} the
mail is either parent or child of re-injection, as explained in
\sectionref{tracking re-injected mail}.

@@

A database is used to store both the rules and the data gleaned by parsing
Postfix log files.  Understanding the database schema is important in
understanding the actions of the parser, and essential to developing
further applications that utilise the data gathered; \sectionref{database
as API} describes how the database schema functions as an \acronym{API}.

@@

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected an \acronym{SMTP} command;
the individual fields will be described in \sectionref{connections table}.
The results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the database schema picture}, to complement the
in-depth descriptions of each table.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection because each result references the rule that
created it.  No ambiguity or confusion is possible about which rule
resulted in a particular result, eliminating one potential source of
confusion.

A clear, comprehensible schema is essential when using the data extracted
from log lines; it is more important when using the data than when storing
it, because storing the data is a write-once operation, whereas utilising
the data requires frequent searching, sorting, and manipulation of the data
to produce customised reports or statistics.

@@

The database populated by \parsername{} provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions in a library, it gives the same benefits as any other
\acronym{API}: it provides a stable interface between the user and the
creator of the data, allowing code on either side of the interface to be
changed without adverse effects, as long as the interface is adhered to.
Programs that use the database can range from the simple examples in
\sectionref{motivation} to far more complex data mining tools.

@@

    \item Databases provide complex querying and sorting functionality for
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, that can be used for ad hoc queries with minimal
        investment of time.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = "boss@example.com"
        AND action = "DELIVERY_REJECTED"
    SEND PANIC EMAIL TO "postmaster@example.com";
\end{verbatim}

            \item Other tables can be added to the database, e.g.\ to cache
                historical or computed data, or to link in data from other
                sources.

@@

Storing the results in a database will also increase the efficiency of
using those results, because the log files need only be parsed once rather
than each time the data is used; indeed the results may be used by someone
with no access to the original log files.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here along with the other
tables in the database.  Rules are created by the user, not the parser, and
will not be modified by \parsername{} (except for the hits and hits\_total
fields).  Rules classify the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
take for each recognised log line.

Each rule some or all of these fields; all fields are required unless
otherwise stated.

@@

    \item [id] A unique identifier for each rule that other tables can use
        when referring to a specific rule.

@@

    \item [description] Something must have occurred to cause Postfix to
        output each log line; this field should describe the event causing
        the log lines this rule recognises.

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  This field is valid only for rules
        that recognise rejection log lines, i.e.\ rules that have an action
        of \action{DELIVERY\_REJECTED}.

    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
        lines the rule applies to; see \sectionref{rule conditions in
        implementation} for full details of how this attribute is used.

@@

    \item [connection\_data] Sometimes rules need to provide data that is
        not present in the log line: e.g.\ setting \texttt{client\_ip} when
        a mail is being delivered to another server.  Any field in the
        connections table can be set in this way.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements.  Commas
        and semi-colons cannot be escaped and thus cannot be included in
        data, this feature is intended for use with small amounts of data,
        so dealing with escape sequences was deemed unnecessary.  This
        field is optional.

@@

    \item [action] The action that will be invoked when this rule
        recognises a log line; a full list of actions and the parameters
        they are invoked with can be found in \sectionref{actions in detail
        in implementation}.

@@

    \item [hits\_total] The total number of hits for this rule over all
        runs of the parser; hits is reset to zero each time the parser is
        run, but hits\_total is not.  This field will be set by the parser
        rather than the rule author.

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules, as
        described in \sectionref{rules in architecture}.  This field is
        optional.

@@

Every accepted mail and every connection that rejected an \acronym{SMTP}
command will have a single entry in the connections table containing all of
the following fields:

@@

    \item [server\_ip] The IP address of the server: the local address when
        receiving mail, the remote address when sending mail.

@@

    \item [client\_ip] The client IP address: the remote address when
        receiving mail, the local address when sending mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \acronym{SMTP} command,
        but it is quite simple to rectify this as described in
        \sectionref{logging helo}.

@@

Every recognised log line will have a row in the results table, associated
with a single connection; a single connection will have at least one result
associated with it, but will usually have several more, and may have
hundreds.

@@

    \item [rule\_id] The id of the rule in the rules table that recognised
        the log line and created this result.

@@

    \item [warning] Administrators can configure Postfix to log a warning
        instead of enforcing a restriction that would reject an
        \acronym{SMTP} command --- a mechanism that is quite useful for
        testing new restrictions.  This field will be false for a real
        rejection, or true if the log line was a warning.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not
        \action{DELIVERY\_REJECTED}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line representing a rejection or final delivery; results whose log
        line did not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code,
        e.g.\ when Postfix delivers to a command; in those cases the
        \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.

@@

    \item [sender] The sender's email address.  This may change during a
        connection if the client uses different sender addresses for
        multiple rejected delivery attempts; however there should not be
        different sender addresses in the results for one accepted mail.
        Mails sent over the same connection can be distinguished by their
        queueids.

@@

    \item [size] The size of the mail, present for delivered mails only.

@@

    \item [data] A field available to store a piece of captured data that
        does not have a specific field, e.g.\ the rejection message from a
        \acronym{DNSBL}\@.

    \item [timestamp] The time at which the log line was logged.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables are used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

When the framework loads the ruleset, it verifies each of the rules,
checking:

@@

    \item That the regex is valid.  The regex will also be compiled for
        efficiency (\sectionref{Caching compiled regexes}).

@@

State tables from a previous \parsername{} run, if any, will be loaded now;
the framework supports saving state at any time, without adversely
affecting the parsing process.  The need to track re-injected mail
(\sectionref{Re-injected mails}) complicates the process of saving state,
as the relationships between mails must be maintained; loading state is
also complicated by dealing with aborted delivery attempts
(\sectionref{aborted delivery attempts}), because a separate set of
relationships between connections and mails must be re-created when the
state tables are restored.

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and warning if the log line is
unrecognised.  The rules are sorted for increased efficiency; see
\sectionref{rule ordering for efficiency} for details.  The repetitive
nature of log files gives them high compression ratios; the framework uses
a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
compressed log files; this saves users the trouble of uncompressing them to
a temporary file before parsing begins.  When used interactively, the
framework also displays a progress bar to show how far parsing has
progressed through the log file and how long the remainder of the parsing
process is likely to take.  The progress bar is not as reliable when
parsing compressed log files, because a compressed block may uncompress to
a variable number of log lines; variation between the recognition and
processing time of individual log lines also affects its accuracy, but
overall the progress bar is a useful addition.  \parsername{} is primarily
intended for parsing complete log files, but with minor modifications it
could be used to parse a live log file, periodically checking it for new
log lines; this could be very useful for programs that work best with up to
date data, e.g.\ a program for monitoring the health of the mail system.

@@

    \item The number of rules used when recognising log lines
        (\graphref{Mean number of rules used per log line}); the framework
        may need to try multiple rules for a given log line before it finds
        one that recognises it.

@@

    \item The framework can record which rule recognised each log line, and
        then on a subsequent run consult that information so that it uses
        the correct rule for each log line.  This gives the best possible
        running time, because only one rule is used for each log line.  A
        variation of this is to use the correct rule last, to get the worst
        possible running time.  \sectionref{perfect rule ordering}
        discusses this in detail.

    \item Each regex is compiled and the result cached when the rules are
        loaded; this can be disabled and the regex compiled every time it
        is used when recognising log lines.  The effect of this
        optimisation is shown in \sectionref{Caching compiled regexes}.

    \item Normally when a log line has been recognised the framework
        invokes the action specified by the rule.  Invoking the action can
        be skipped if desired, so the timing information shows how long is
        spent recognising log lines only; this time can be subtracted from
        the time taken by a normal run to show how long is spent processing
        log lines.  This data is analysed in \sectionref{recognising vs
        processing}.

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser; the evaluation
        chapter measures the speed of \parsername{}, not the speed of the
        database or the disks it resides on.  The effect on parsing time of
        storing results in the database is described at the beginning of
        \sectionref{Evaluation}.

@@

    \item Each result can be extended with extra debugging information,
        which is useful when a warning dumps a mail's data for inspection.
        The information that is added is: the log line's timestamp in human
        readable form; the entire log line; the name of the log file and
        the number of the log line within it.  This information is not
        stored in the database.

    \item Every time a log line is recognised, the recognising rule's regex
        and the log line can be printed, so that the user can verify that
        the correct rule recognised each log line.

@@

If the log line has enough information to identify the correct connection
or mail, each action will save all the data captured by the recognising
rule's regex.  Each action is passed the same arguments:

>>>>

This chapter documents the implementation of \parsername{}, a parsed based
on the architecture described in \sectionref{parser architecture}.  Any
design may look good when examined in the abstract, but the real test of
the design comes with the first concrete implementation; only then do any
practical difficulties come to light.  Implementing this architecture was
relatively straightforward: the difficulties came not from the architecture
and design, but from anomalies in the log files and Postfix's behaviour, as
described in \sectionref{complications}.  \parsername{} successfully deals
with all the difficulties that were discovered during its development, but
further difficulties may arise during future usage; unfortunately solving
all potential difficulties that may arise is an intractable task, but the
descriptions of the solutions developed thus far should help if someone
needs to develop a solution of their own.

\parsername{} deals with all the eccentricities and oddities of parsing
Postfix log files, presenting the resulting data in a normalised, simple to
use representation.  Unfortunately, dealing with the complications that
arise sometimes requires the parser to discard log lines (e.g.\
\sectionref{aborted delivery attempts}), and, less frequently, to discard a
data structure (e.g.\ \sectionref{timeouts during data phase}).
\parsername{} can parse log files from Postfix version 2.2 through to
version 2.5, and should be capable of parsing log files from later versions
with only minor modifications or an updated ruleset.  The parser's task is
to follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must handle multiple mails in
parallel, each potentially at a different stage in its journey, without any
interference between mails --- except in the minority of cases where
intra-mail interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

This chapter begins with the assumptions under which \parsername{} was
designed and written, followed by a flowchart showing the most common paths
taken through Postfix and \parsername{}, with a description of the stages
and stage transitions.  The second topic is the \acronym{SQL} database that
provides storage for the parser: any future work that analyses gathered
data will do so using the database, so the database schema acts as an
\acronym{API}.  A diagram of the database schema is provided, plus
documentation for every table and field.

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users of
\parsername{}; that section finishes with the features that are important
for this thesis but not strictly necessary for parsing: the performance
data collected by the framework, the optimisations that can be disabled to
show their effect, and the debugging options it provides.  The second
component of the architecture is the actions, starting with a graph of how
often each action is specified by rules, a description of why some actions
are more popular than others, and how this popularity has no influence on
how often an action is invoked.  Every action that is part of \parsername{}
is documented, and the actions section concludes with a description of the
process of adding a new action.  Rules are the final component of the
architecture, and also the most visible to advanced users, e.g.\ systems
administrators, because it is likely that they will need to add their own
rules to recognise their log lines.  An example rule used by \parsername{}
is examined, with every field clearly documented; that is followed by a
description of adding new rules, and how to determine the value of each of
the fields.  \parsername{} provides a utility to create regexes from
unrecognised log lines; the algorithm it uses is documented, including the
differences between it and the original algorithm it is based on.  The
rules section finishes with a discussion of how \parsername{} uses rule
conditions and overlapping rules, plus a description of the regex snippets
provided to ease the process of writing regexes.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run mail server it is
        extremely unlikely that any of these problems will arise, though
        the likelihood increases when suffering from a deluge of spam or a
        mail loop.  When parsing individual log files in isolation, it is
        highly likely that some mails will have log lines in previous log
        files, and others will have log lines in subsequent log files; to
        alleviate this problem \parsername{} supports saving and loading
        its state tables, so they can be preserved between log files.

@@

In some ways this task is similar to reverse engineering or replicating a
black box system based solely on its inputs and outputs.  Thus far,
analysis of the log files has been enough to reconstruct Postfix's
behaviour, but for other programs the techniques described in
\cite{black-box-error-reporting} may be useful.  Some advantages come from
treating Postfix as a black box during parser development:

@@

    \item The parser acts as a second source of information about Postfix's
        operation, based on empirical evidence from log files.  A separate
        project could compare the empirical knowledge inherent in
        \parsername{} with Postfix's documentation and source code to see
        how closely the two agree.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; decision boxes and the
difficulties described in \sectionref{complications} are excluded for the
sake of clarity.  The flow chart is intended to be a graphical overview of
how a mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and state transitions shown in the flow chart will be explained
later in this section.

@@

    \item This is followed by rejection of one or more mail delivery attempts
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; failure of the remote client's connection
        \flowchart{DELIVERY\_ERROR}{6}; or some random interleaving of two
        or more of the above.

    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
        normally or with an error.  If Postfix has rejected any
        \acronym{SMTP} commands the data gathered from those rejections
        will be saved to the database; if there were no rejections there
        will not be any data to save.  Any mails accepted will already have
        been cloned so their data is in another data structure, and will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether by deletion, bouncing, expiry, local delivery, or remote
delivery.  All five are handled in the same way:

@@

    \item Sometimes mail is re-injected for forwarding and the child mail
        needs to be tracked with the parent mail \flowchart{TRACK}{9}; the
        handling of re-injected mails is described in \sectionref{tracking
        re-injected mail}.

@@

    \item The mail is removed from the Postfix queue.  This is the last log
        line for this particular mail, though it may be indirectly referred
        to if it was re-injected.  The mail is cleaned up and entered in
        the database, then deleted from the state tables
        \flowchart{COMMIT}{11}.

@@

It should be emphasised that the sequence above happens whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (because of a mail loop, delivery failure, or five day timeout), or
deleted by the administrator, \textit{unless\/} the mail is either parent
or child of re-injection, as explained in \sectionref{tracking re-injected
mail}.

@@

A database is used to store both the rules and the data gleaned by parsing
Postfix log files.  Understanding the database schema is important in
understanding the actions of the parser, and essential to developing
further applications that utilise the data; \sectionref{database as API}
describes how the database schema functions as an \acronym{API}.

@@

The data saved from parsing the log files is divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected a delivery attempt; the
individual fields will be described in \sectionref{connections table}.  The
results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the database schema picture}, to complement the
in-depth descriptions of each table.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection because each result references the rule that
created it.  No ambiguity is possible about which rule resulted in a
particular result, eliminating one potential source of confusion.

A clear, comprehensible schema is essential when using the data extracted
from log lines; it is more important when using the data than when storing
it, because storing the data is a once-off operation, whereas utilising
the data requires frequent searching, sorting, and manipulation of the data
to produce customised reports or statistics.

@@

The database populated by \parsername{} provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions in a library, it provides the same benefits as any other
\acronym{API}: a stable interface between the user and the creator of the
data, allowing code on either side of the interface to be changed without
adverse effects, as long as the interface is adhered to.  Programs that use
the database can range from the simple examples in \sectionref{motivation}
to far more complex data mining tools and machine learning algorithms.

@@

    \item Databases provide complex querying and sorting functionality for
        the user without requiring large amounts of programming.  All
        databases have one or more programs, of varying complexity and
        sophistication, that can be used for ad hoc queries with minimal
        investment of time.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = "boss@example.com"
        AND rule_id = rules.id
        AND rules.action = "DELIVERY_REJECTED"
    SEND PANIC EMAIL TO "postmaster@example.com";
\end{verbatim}

            \item Other tables can be added to the database, e.g.\ to cache
                historical or computed data, or to incorporate data from
                other sources.

@@

Storing the results in a database will also increase the efficiency of
using those results, because the log files only need to be parsed once
rather than each time the data is used; indeed the database may be used by
someone with no access to the original log files.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here alongside the other
tables in the database.  Rules are created by the user, and will not be
modified by \parsername{}, except when it updates the hits and hits\_total
fields.  Rules recognise the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
invoke for each recognised log line.

Each rule must provide values for most of the fields below; all fields are
required unless otherwise stated in the description.

@@

    \item [id] A unique identifier that other tables can use when referring
        to a specific rule.

@@

    \item [description] This field should describe the event causing the
        log lines this rule recognises, e.g.\ ``Mail has been delivered to
        the LDA (typically procmail)''.

    \item [restriction\_name] The name of the Postfix restriction that
        caused the rejection of the mail delivery attempt.  This field is
        valid only for rules that recognise rejection log lines, i.e.\
        rules that have an action of \action{DELIVERY\_REJECTED}.

    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
        lines the rule recognises; see \sectionref{rule conditions in
        implementation} for full details of how this attribute is used.

@@

    \item [connection\_data] Sometimes rules need to provide data that is
        not present in the log line, e.g.\ setting \texttt{client\_ip} when
        a mail is being delivered to another server; any field in the
        connections table can be set in this way.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements.  Commas
        and semi-colons cannot be escaped and thus cannot be included in
        data, because this feature is intended for use with small amounts
        of data and dealing with escape sequences was deemed unnecessary.
        This field is optional.

@@

    \item [action] The action to be invoked when this rule recognises a log
        line; a full list of actions and the parameters they are invoked
        with can be found in \sectionref{actions in detail in
        implementation}.

@@

    \item [hits\_total] The total number of log lines recognised by this
        rule over all runs of the parser; hits is reset to zero each time
        the parser is run, but hits\_total is not.  This field will be set
        by the parser rather than the rule author.

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted by the parser, priority overrides hits.
        This allows more specific rules to take precedence over more
        general rules, as described in \sectionref{rules in architecture}.
        This field is optional.

@@

Every accepted mail and every connection that rejected a mail delivery
attempt will have a single entry in the connections table containing all of
the following below.  For an incoming connection, the client is the remote
machine, and the server is the local machine; for outbound mail delivery
attempts, the roles are reversed.

@@

    \item [server\_ip] The IP address of the server.

@@

    \item [client\_ip] The client IP address.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects a mail delivery attempt, but
        it is quite simple to rectify this as described in
        \sectionref{logging helo}.

@@

Every recognised log line will have a row in the results table, associated
with a single connection; a single connection will have at least one result
associated with it, but will usually have several, and may have hundreds.

@@

    \item [rule\_id] The id of the rule in the rules table that recognised
        the log line.

@@

    \item [warning] Administrators can configure Postfix to log a warning
        instead of enforcing a restriction that would reject a mail
        delivery attempt --- a mechanism that is quite useful for testing
        new restrictions.  This field will be false for a real rejection,
        or true if the log line was a warning.  This field should be
        ignored if the result is not a rejection, i.e.\ the action field of
        the associated rule is not \action{DELIVERY\_REJECTED}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line representing a rejection or final delivery; results whose log
        line did not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code,
        e.g.\ when Postfix delivers to a user's mailbox; in those cases the
        \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.

@@

    \item [sender] The sender's email address.  This may change during a
        connection if the client uses different sender addresses for
        multiple rejected delivery attempts; however there will not be
        different sender addresses in the results for one accepted mail.

@@

    \item [size] The size of the mail, only available for delivered mails.

@@

    \item [data] A field available to store a piece of captured data that
        does not have its own specific field, e.g.\ the rejection message
        from a \acronym{DNSBL}\@.

    \item [timestamp] The log line's timestamp.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up several state tables that will later be used by actions.
Data about the connections and mails being processed is held in
\texttt{connections} and \texttt{queueids} respectively.  The other state
tables are used when working around complications in parsing:
\texttt{timeout\_queueids} is used when dealing with connections that time
out during the DATA phase (\sectionref{timeouts during data phase});
\texttt{bounce\_queueids} is part of the solution to bounce notification
mails being delivered before their creation is logged (\sectionref{Bounce
notification mails delivered before their creation is logged}); and
\texttt{postsuper\_deleted\_queueids} caches information about mails that
were recently deleted by the administrator so that subsequent log lines
processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

The framework verifies each of the rules when it loads the ruleset,
checking:

@@

    \item That the regex is valid.  The regex will have regex components
        expanded (see \sectionref{regex components}), and will also be
        compiled for efficiency (\sectionref{Caching compiled regexes}).

@@

State tables from a previous \parsername{} run, if any, will be loaded now;
the framework supports saving state at any time, without adversely
affecting the parsing process.  The need to track re-injected mail
(\sectionref{Re-injected mails}) complicates the process of saving state,
because the relationships between mails must be maintained; loading state
is also complicated by dealing with aborted delivery attempts
(\sectionref{aborted delivery attempts}), because a separate set of
relationships between connections and mails must be re-created when the
state tables are restored.  The last step in loading the ruleset is to sort
the rules as described in \sectionref{rule ordering for efficiency}.

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and finally warning if the log
line is unrecognised.  The repetitive nature of log files gives them high
compression ratios; the framework uses a Perl module named
\texttt{IO::Uncompress::AnyUncompress} to read compressed log files, saving
users the trouble of uncompressing them to a temporary file before parsing
begins.  When used interactively, the framework displays a progress bar to
show how far parsing has progressed through the log file and how long the
remainder of the parsing process is likely to take.  The progress bar is
not as accurate when parsing compressed log files, because a compressed
block uncompresses to a variable number of log lines; variation between the
recognition and processing time of individual log lines also affects its
accuracy, but overall the progress bar is a useful addition.  \parsername{}
is primarily intended for parsing complete log files, but with minor
modifications it could be used to parse a live log file, periodically
checking it for new log lines; this could be very useful for programs that
work best with up to date data, e.g.\ a program for monitoring the health
of the mail system, or graphs showing activity over the last five minutes.

@@

    \item The number of rules used when recognising log lines
        (\graphref{Mean number of rules used per log line}); the framework
        may need to try multiple rules for each log line before it finds
        one that recognises it.

@@

    \item The framework can record which rule recognised each log line, and
        then on a subsequent run consult that information so that it uses
        the correct rule for each log line.  This gives the best possible
        running time, because only one rule is used to recognise each log
        line, as discussed in \sectionref{perfect rule ordering}.

    \item Each regex is compiled and the result cached when the ruleset is
        loaded; this optimisation can be disabled and the regex compiled
        every time it is used when recognising log lines.  The effect of
        this optimisation is shown in \sectionref{Caching compiled
        regexes}.

    \item Normally, when a log line has been recognised the framework
        invokes the action specified by the rule.  Invoking the action can
        be skipped if desired, so the timing information shows how long is
        spent recognising log lines only; this time can be subtracted from
        the time taken by a normal run to show how long is spent processing
        log lines.  This data is analysed in \sectionref{recognising vs
        processing}.

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser; the evaluation
        chapter measures the speed of \parsername{}, not the speed of the
        database or the disks it resides on.  The effect on parsing time of
        storing results in the database is described at the beginning of
        \sectionref{parser efficiency}.

@@

    \item Each result can be extended with extra debugging information,
        which is useful when a warning dumps a mail's data for inspection.
        The extra information added is: the log line's timestamp in human
        readable form; the entire log line; the name of the log file and
        the number of the log line within it.  This information is not
        stored in the database.

    \item Every time a log line is recognised, the recognising rule's regex
        and the log line can be printed, so that the user can verify that
        the correct rule recognised each log line.  This is equivalent to
        setting every rule's debug field to true.

@@

If the log line has enough information to identify the correct connection
or mail, each action will save all the data captured by the recognising
rule's regex; usually, log lines that lack identifying information will be
processed by the \action{UNINTERESTING} action.  Each action is passed the
same arguments:
